Predicting the conformational preferences of proteins using a physics-based free energy method

### 2013-03-06 Paper/paper.tex
Made changes to introduction

Protein conformation and associated conformational changes are important in biology. The mechanism
of enzymes, the function of structural proteins, and the binding of proteins to DNA or other
proteins all depend on protein conformations and the relative free energies of different
conformational states. Proteins can undergo conformational changes for a variety of reasons,
including ligand binding, change of pH or other conditions \cite{Meirovitch2007}, \cite{Chipot2007},
\cite{Jorgensen2004}. The process of protein folding also depends on the free energies of different
conformational states. The native structure is at a free energy minimum \cite{Anfinsen1973}, while
the kinetics are determined by the free energies of various on- and off-pathway intermediates.

However, the theoretical calculation of conformational free energy changes is difficult when the
states of interest are very different \cite{Meirovitch2007}. In these cases, the timescales involved
in such conformational transitions may be too long to simulate directly. For example, in protein
folding the free energy landscape is rough \cite{Dill1997}, \cite{Dill2008}, and the molecule can
become trapped in an energy well during conformational sampling. Even with advanced methods like
replica exchange molecular dynamics, the three dimensional structure may be trapped in a local
minima for a considerable time. Sometimes this problem can be circumvented by defining a pathway or
reaction coordinate connecting the states of interest. The free energy along this reaction
coordinate can then be determined using methods such as umbrella sampling \cite{Torrie1977},
targeted MD \cite{}, or thermodynamic integration \cite{Tironi1994}. However, accurately sampling
this reaction coordinate may still be difficult and require a large computationl effort. In the case
of protein folding, it is often difficult to even devise a meaningful reaction coordinate that could
connect the unfolded and folded states efficiently \cite{}.

The calculation of protein conformational free energy has been successfully attempted by a number of
groups \cite{Meirovitch2007,Ytreberg2006,Zheng2008}. In recent years, methods such as the reference
system method \cite{Ytreberg2006}, decativated morphing \cite{Park2008}, orthogonal space random
walk \cite{Zheng2008}, and the confinement method \cite{Tyka2006,Cecchini2009} have been used to
calculate free energy differences.

In this work we explore confinement method\cite{Tyka2006,Cecchini2009}, which relies on the fact
that the free energy difference is a state function and thus is independent of the sampling path.
This approach uses a thermodynamic cycle \ref{fig:method} where first protein conformational entropy
of the two end-states is frozen out by applying a series of restraints to the system. Next, the free
energy difference between the two highly-restrained harmocic systems is obtained using a normal mode
calculation.

Previously, this method was applied to calculate the conformational free energy of a 16 amino acid
residue peptide, known as BHP.  We have first reproduced that known result. Next, we have applied
this method to larger proteins in two different case studies. In the first, we attempt to pick the
most native like structures from a set of models produced during the Critical Assessment of
Structure Prediction (CASP) \cite{} experiment. In the second case study, we examine the
conformational preferences of a series of chameleon proteins designed by Orban and co-workers
\cite{} that can switch between two completely different folds with only small changes in amino acid
sequence.

>>>>

Protein structure and conformational changes are of central importance in biochemistry. The
populations of different conformational states are determined by their free energies. The ability to
compute these free energies would allow us to understand and make predictions about a wide variety
of biochemical processes---including the mechanism of enzymes, the function of structural proteins,
binding of proteins to other proteins or DNA, and protein folding.

Computer simulations have proven to be a powerful tool for studying biomolecular systems at atomic
detail [ref]. However, calculating the free energies for conformational changes remains a
challenging problem \cite{Meirovitch2007}, primarily because the timescales for such conformational
transitions may be too long to simulate directly. For example, in protein folding the free energy
landscape is very rough \cite{Dill1997,Dill2008}, which limits conformational sampling and leads to
large uncertainties in the relative populations of different states.

Even with advanced methods like replica exchange molecular dynamics [ref], the molecule may be
become trapped in a local minima. Sometimes this problem may be circumvented by defining a pathway
or reaction coordinate connecting the states of interest. The free energy along this reaction
coordinate can then be determined using methods such as umbrella sampling \cite{Torrie1977},
targeted MD \cite{}, or thermodynamic integration \cite{Tironi1994}. However, accurately sampling
this reaction coordinate may still be difficult and require a large computational effort. In the
case of protein folding, it is often difficult to even devise a meaningful reaction coordinate that
could connect the unfolded and folded states efficiently \cite{}.

The calculation of protein conformational free energy has been successfully attempted by a number of
groups \cite{Meirovitch2007,Ytreberg2006,Zheng2008}. In recent years, methods such as the reference
system method \cite{Ytreberg2006}, deactivated morphing \cite{Park2008}, orthogonal space random
walk \cite{Zheng2008}, and the confinement method \cite{Tyka2006,Cecchini2009} have been used to
calculate free energy differences.

In this work we explore the confinement method\cite{Tyka2006,Cecchini2009}, which relies on the fact
that the free energy difference is a state function and thus is independent of the sampling path.
This allows us to use the alchemical thermodynamic cycle shown in Figure \ref{fig:method}. First,
the two end states are slowly ``frozen'' into a very tight, harmonically restrained ensemble and the
free energies for these two processes are calculated. Next, the free energy difference between the
two restrained ensembles is calculated using normal mode analysis. This allows us to calculate the
free energy difference between the two end states without needing to define a physical path or
reaction coordinate connecting the two states.

The confinement has previously been applied to several small proteins [refs]. Here, we use the power
of GPU computing (using graphic cards to run calculations [cite amber gpu]), which allows us to
extend the method to larger systems while maintaining reasonable computation times. We first
demonstrate that our implementation reproduces the previously reported results for small peptides.
Next, we have applied this method to larger proteins in two different case studies. In the first, we
attempt to pick the most native-like structures from a set of models produced during the Critical
Assessment of Structure Prediction (CASP) \cite{} experiment. In the second case study, we examine
the conformational preferences of a series of chameleon proteins designed by Orban and co-workers
\cite{} that can switch between two completely different folds with only small changes in amino acid
sequence.
